---
title: "Navigating Through Crisis: The Influence of COVID-19 on Canada’s Mental Health Problem"
subtitle: "My subtitle if needed"
author: 
  - Zijun Meng
thanks: "Code and data are available at: https://github.com/FrankMengZJ/canada_mental_health/"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(arrow)
library(here)
library(knitr)
age <- read_parquet(file = here("data/analysis_data/age.parquet"))
mental_health_indicators <- read_parquet(file = here( "data/analysis_data/mental_health_indicators.parquet"))
province <- read_parquet(file = here("data/analysis_data/province.parquet"))
sc <- read_parquet(file = here("data/analysis_data/sociodemographic_characteristics.parquet"))
```

# Introduction

The COVID-19 pandemic has spotlighted mental health as a critical component of public health, magnifying existing challenges and uncovering gaps within healthcare systems globally. In Canada, the impact of the pandemic on mental health has been profound, exacerbating disparities and highlighting the urgency of effective mental health care. The Centre for Addiction and Mental Health (CAMH) has identified a significant increase in mental health problems among Canadians, noting that the pandemic has led to unprecedented levels of anxiety and depression (CAMH, 2020). Additionally, the Public Health Agency of Canada reports on the elevated symptoms of anxiety and depression during the pandemic, further underscoring the need for comprehensive data analysis on this issue (Public Health Agency of Canada, 2021).

This paper aims to provide a detailed analysis of the changing situation of mental health in Canada by using four datasets spanning from 2002 to 2023. The focus is on perceived mental health statuses across various demographic segments, including age, gender, and province. The primary objective is to understand how perceptions of mental health have evolved over the past two decades and how these perceptions have been impacted by the COVID-19 pandemic.

The study addresses the gap in longitudinal mental health research, particularly in the analysis of sociodemographic influences during a major global crisis. Through the comparison of mental health indicators before and during the pandemic, this research provides insights into the shifts in mental health perceptions and the effectiveness of the responses to these changes. Initial findings suggest notable disparities in mental health impacts across different demographic groups, indicating the need for targeted interventions.

The significance of this research lies in its contribution to the evidence base informing mental health policy and practice. By identifying the demographic factors associated with varying mental health outcomes, the study aids policymakers and healthcare providers in developing tailored strategies to address the unique challenges faced by diverse groups within Canada. 
*TBD*



# Data {#sec-data}

This study uses four carefully curated datasets from Statistics Canada, each providing valuable insights into the mental health perceptions of Canadians. These datasets are pivotal for understanding the evolution of mental health across different demographics and geographical regions from 2002 to 2023.

## Overview of Datasets

The data for this research come from national surveys conducted by Statistics Canada, specifically designed to monitor various aspects of mental health across the country. Each dataset targets different demographic variables, such as age, gender, and province, and covers distinct periods critical for analyzing long-term trends and the immediate effects of the COVID-19 pandemic.


### Mental Health Indicators Dataset
Data from 2002 and 2012 in the Mental Health Indicators Dataset were derived from national health databases, capturing key statistics on the prevalence of mental disorders and utilization of mental health services. The information is gathered through comprehensive health surveys and administrative records, providing robust measures of mental health status and service outcomes.

@tbl-1 shows the first 5 rows in the dataset, REF_DATE stands for the time that the data was gathered, GEO denotes the geographical ,

```{r}
#| label: tbl-1
#| tbl-cap: First 5 Rows in Mental Health Indicators Dataset 
#| echo: false

kable(head(mental_health_indicators,5))
```

### Perceived Mental Health by Age Group
The Perceived Mental Health by Age Group dataset spans from 2015 to 2022 and utilizes a consistent methodological framework to assess self-reported mental health. The data collection process involves standardized questionnaires distributed to a representative sample of Canadian adults across various age groups. This approach ensures that the data reflect a broad spectrum of the population, excluding those in full-time residence, on Indigenous reserves, or non-permanent residents, thus focusing on a generalizable sample of the Canadian populace.

### Perceived Mental Health by Gender and Sociodemographic Characteristics
The data for 2021 to 2023 from this dataset is sourced from the Canadian Community Health Survey (CCHS), which meticulously gathers data on mental health perceptions influenced by gender and other sociodemographic factors such as income, education, and employment. This survey selects participants through a stratified sampling design aimed at capturing diverse sociodemographic landscapes across Canada, enhancing the dataset's relevance for analyzing intersectional impacts on mental health.

### Detailed Measurement Discussion

Both the CCHS and other Statistics Canada surveys employ rigorous data verification and validation processes. Quality control measures include logical checks during data collection and detailed follow-ups for non-responses. Post-stratification weighting adjusts for any discrepancies due to non-response rates, and seasonal adjustments are made to account for the timing of data collection. These methodologies ensure that the data are robust, reliable, and accurately reflect the Canadian population’s health status.

The CCHS's use of health-specific, standardized questionnaires is designed to capture nuanced information on health determinants and outcomes, which is critical for public health research and intervention planning. Conversely, Statistics Canada’s broader surveys provide comprehensive data across a wider range of topics, offering valuable insights into the interplay between health and other societal factors.

### Comparison of Data Collection Specifics

CCHS is highly adaptable with its questionnaire content, allowing for the inclusion of regional health issues of specific interest, which is crucial for targeted public health interventions. This adaptability in data collection, combined with its detailed focus on health, makes CCHS uniquely suited for studies like ours that require specific health-related insights.

In summary, the methodologies used by the Canadian Community Health Survey and other Statistics Canada surveys are both robust and tailored to meet different informational needs. Understanding these methodologies helps in accurately interpreting the data collected on mental health perceptions, ensuring that research findings are based on reliable and representative data.

# Results

Our results are summarized in .

## Overview


```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-trend1
#| fig-cap: Perceived mental health fair or poor in Canada, 2002-2022

mhi <- filter(mental_health_indicators, Indicators == "Perceived mental health, fair or poor" & Age == "Total, 15 years and over" & Sex == "Both sexes" & GEO == "Canada")

ag <- filter(age, Indicators == "Perceived mental health, fair or poor" & Age == "Total, 12 years and over" & Sex == "Both sexes" & GEO == "Canada (excluding territories)")

temp <-
  rbind(select(mhi,REF_DATE,VALUE),select(ag,REF_DATE,VALUE))

ggplot(temp, aes(x=REF_DATE,y=VALUE))+
  geom_line()+
  geom_point()+
  labs(x = "Year", y = "Percentage") +
  theme_linedraw()

```


```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-trend2
#| fig-cap: Perceived mental health fair or poor in Canada

temp <- filter(sc, Indicators == "Fair or poor perceived mental health" & Sociodemographic_characteristics  == "Total, 15 years and over" & Gender == "Total, all persons" & GEO == "Canada (excluding territories)")


ggplot(temp, aes(x=REF_DATE,y=VALUE))+
  geom_line(aes(group=1))+
  geom_line()+
  geom_point()+
  labs(x = "Year", y = "Percentage") +
  theme_linedraw()

```
## By Province

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-trend3
#| fig-cap: Perceived mental health fair or poor in Canada

temp <- filter(age, Indicators == "Perceived mental health, fair or poor" & Sex == "Both sexes", Age =="Total, 12 years and over")

temp <-
  temp|>
  select(REF_DATE,GEO,VALUE)

temp1 <-
  group_by(temp, GEO) |>
  summarise(variance = var(VALUE))

ggplot(temp, aes(x= REF_DATE, y= VALUE, group = GEO, color = GEO))+
  geom_line()+
  theme_linedraw()

```


```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-trend4
#| fig-cap: Perceived mental health fair or poor in Canada

temp <- filter(province, Indicators == "Fair or poor perceived mental health" & Gender == "Total, all persons")

temp <-
  temp|>
  select(REF_DATE,GEO,VALUE)

temp1 <-
  group_by(temp, GEO) |>
  summarise(variance = var(VALUE))

ggplot(temp, aes(x= REF_DATE, y= VALUE, group = GEO, color = GEO))+
  geom_line()+
  theme_linedraw()

```

## By Gender

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-trend4
#| fig-cap: Perceived mental health fair or poor in Canada

temp <- filter(age, Indicators == "Perceived mental health, fair or poor" & GEO == "Canada (excluding territories)"& Age == "Total, 12 years and over")


temp1 <-
  group_by(temp, Sex)

colors <- c("Males" = "blue", "Females" = "green4","Both sexes" = "red")

ggplot(temp, aes(x= REF_DATE, y= VALUE, group = Sex, color = Sex))+
  geom_line()+
  geom_point()+
  scale_color_manual(values = colors) +
  theme_linedraw()

```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-trend2
#| fig-cap: Perceived mental health fair or poor in Canada

temp <- filter(sc, Indicators == "Fair or poor perceived mental health" & Sociodemographic_characteristics  == "Total, 15 years and over" & GEO == "Canada (excluding territories)")

colors <- c("Men" = "blue", "Women" = "green4","Total, all persons" = "red")

ggplot(temp, aes(x=REF_DATE,y=VALUE, group = Gender, color = Gender))+
  geom_line()+
  geom_point()+
  labs(x = "Year", y = "Percentage") +
  scale_color_manual(values = colors) +
  theme_linedraw()

```
## By Age
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-trend4
#| fig-cap: Perceived mental health fair or poor in Canada

temp <- filter(age, Indicators == "Perceived mental health, fair or poor" & GEO == "Canada (excluding territories)" & Sex == "Both sexes")


ggplot(temp, aes(x= REF_DATE, y= VALUE, group = Age, color = Age))+
  geom_line()+
  geom_point()+
  theme_linedraw()

```


```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-trend2
#| fig-cap: Perceived mental health fair or poor in Canada

temp <- filter(sc, Indicators == "Fair or poor perceived mental health" & Gender  == "Total, all persons" & GEO == "Canada (excluding territories)" & grepl('years',Sociodemographic_characteristics)& !grepl("Immigrants",Sociodemographic_characteristics))


ggplot(temp, aes(x=REF_DATE,y=VALUE, group = Sociodemographic_characteristics , color = Sociodemographic_characteristics ))+
  geom_line()+
  geom_point()+
  labs(x = "Year", y = "Percentage") +
  theme_linedraw()

```

## Others

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-trend2
#| fig-cap: Perceived mental health fair or poor in Canada

temp <- filter(sc, Indicators == "Fair or poor perceived mental health")

temp <- drop_na(temp, VALUE)

temp1 <- 
  group_by(temp, GEO, Gender, Sociodemographic_characteristics) |>
  summarise(avg = mean(VALUE))

print(temp1[order(-temp1$avg),],n=10)

```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

# References
