---
title: "Navigating Through Crisis: The Influence of COVID-19 on Canada’s Mental Health Problem"
subtitle: "My subtitle if needed"
author: 
  - Zijun Meng
thanks: "Code and data are available at: https://github.com/FrankMengZJ/canada_mental_health/"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(arrow)
library(here)
library(knitr)
age <- read_parquet(file = here("data/analysis_data/age.parquet"))
mental_health_indicators <- read_parquet(file = here( "data/analysis_data/mental_health_indicators.parquet"))
province <- read_parquet(file = here("data/analysis_data/province.parquet"))
sc <- read_parquet(file = here("data/analysis_data/sociodemographic_characteristics.parquet"))
```

# Introduction

The global COVID-19 pandemic has had a profound impact on public health, particularly in the realm of mental health. In Canada, the challenges brought on by the pandemic have intensified existing mental health issues and revealed significant vulnerabilities within the healthcare system (Government of Canada, 2021) *cite this*. This paper explores the effects of the COVID-19 pandemic on mental health trends in Canada, with a particular focus on the shifts in prevalence and the disparities in healthcare access observed across different sociodemographic groups during this period.

Research on the long-term mental health trends during the COVID-19 pandemic is sparse, particularly studies that provide a comprehensive analysis across various sociodemographic groups. Most existing studies have concentrated on immediate or short-term impacts within narrowly defined populations. This study aims to bridge this gap by conducting a detailed examination of how factors such as age, gender, and geographic location have influenced mental health outcomes across the broader Canadian population during the pandemic (CAMH, 2020) *cite this*.

The methodology of this paper includes a comparative analysis of data from national health surveys, focusing on the period from 2019 to the latest available data in 2021. This approach allows us to observe both pre-pandemic and during-pandemic mental health indicators, identifying significant trends and changes. Preliminary observations suggest that there are significant disparities in the impact of the pandemic on mental health, which vary widely among different demographic groups.

This research is critical not only for academic purposes but also for policymakers and healthcare providers planning interventions for ongoing and future public health crises. The structure of the paper is as follows: Section 1 introduces the background and literature relevant to the study; Section 2 outlines the methodology and data sources; Section 3 presents the analysis of the findings; Section 4 discusses the implications of these findings for mental health policy and practice in Canada.

# Data {#sec-data}

There are four datasets used for this paper, including the 'Mental health indicator' dataset which contains data related to mental health in 2002 and 2012, 'Perceived mental health, by age group' dataset which data between 2015 and 2022, 'Perceived mental health, by gender and other selected sociodemographic characteristics' and 'Perceived mental health, by gender and province' datasets both contain data between 2021 and 2023, but have different variables.

## Mental Health Indicator Dataset

The mental health indicator dataset records 
```{r}
#| label: tbl-1
#| fig-cap: Mental Health Indicator Dataset 
#| echo: false

kable(head(mental_health_indicators,5))
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

```{=tex}
\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}
```
We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {.unnumbered}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References
