---
title: "Navigating Through Crisis: The Influence of COVID-19 on Canada’s Mental Health Problem"
subtitle: "My subtitle if needed"
author: 
  - Zijun Meng
thanks: "Code and data are available at: https://github.com/FrankMengZJ/canada_mental_health/"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(arrow)
library(here)
library(knitr)
age <- read_parquet(file = here("data/analysis_data/age.parquet"))
mental_health_indicators <- read_parquet(file = here( "data/analysis_data/mental_health_indicators.parquet"))
province <- read_parquet(file = here("data/analysis_data/province.parquet"))
sc <- read_parquet(file = here("data/analysis_data/sociodemographic_characteristics.parquet"))
```

# Introduction

The COVID-19 pandemic has spotlighted mental health as a critical component of public health, magnifying existing challenges and uncovering gaps within healthcare systems globally. In Canada, the impact of the pandemic on mental health has been profound, exacerbating disparities and highlighting the urgency of effective mental health care. The Centre for Addiction and Mental Health (CAMH) has identified a significant increase in mental health problems among Canadians, noting that the pandemic has led to unprecedented levels of anxiety and depression (CAMH, 2020). Additionally, the Public Health Agency of Canada reports on the elevated symptoms of anxiety and depression during the pandemic, further underscoring the need for comprehensive data analysis on this issue (Public Health Agency of Canada, 2021).

This paper aims to provide a detailed analysis of the changing situation of mental health in Canada by using four datasets spanning from 2002 to 2023. The focus is on perceived mental health statuses across various demographic segments, including age, gender, and province. The primary objective is to understand how perceptions of mental health have evolved over the past two decades and how these perceptions have been impacted by the COVID-19 pandemic.

The study addresses the gap in longitudinal mental health research, particularly in the analysis of sociodemographic influences during a major global crisis. Through the comparison of mental health indicators before and during the pandemic, this research provides insights into the shifts in mental health perceptions and the effectiveness of the responses to these changes. Initial findings suggest notable disparities in mental health impacts across different demographic groups, indicating the need for targeted interventions.

The significance of this research lies in its contribution to the evidence base informing mental health policy and practice. By identifying the demographic factors associated with varying mental health outcomes, the study aids policymakers and healthcare providers in developing tailored strategies to address the unique challenges faced by diverse groups within Canada. 
*TBD*

# Data {#sec-data}

There are four datasets used for this paper, including the 'Mental health indicator' dataset which contains data related to mental health in 2002 and 2012, 'Perceived mental health, by age group' dataset which data between 2015 and 2022, 'Perceived mental health, by gender and other selected sociodemographic characteristics' and 'Perceived mental health, by gender and province' datasets both contain data between 2021 and 2023, but have different variables.

## Mental Health Indicator Dataset

The mental health indicator dataset records 
```{r}
#| label: tbl-1
#| fig-cap: Mental Health Indicator Dataset 
#| echo: false

kable(head(mental_health_indicators,5))
```

# Data {#sec-data}

This study utilizes four carefully curated datasets from Statistics Canada, each providing valuable insights into the mental health perceptions of Canadians. These datasets are pivotal for understanding the evolution of mental health across different demographics and geographical regions from 2002 to 2023.

## Overview of Datasets

The data for this research come from national surveys conducted by Statistics Canada, specifically designed to monitor various aspects of mental health across the country. Each dataset targets different demographic variables, such as age, gender, and province, and covers distinct periods critical for analyzing long-term trends and the immediate effects of the COVID-19 pandemic.


### Mental Health Indicators Dataset
Data from 2002 and 2012 in the Mental Health Indicators Dataset were derived from national health databases, capturing key statistics on the prevalence of mental disorders and utilization of mental health services. The information is gathered through comprehensive health surveys and administrative records, providing robust measures of mental health status and service outcomes.

### Perceived Mental Health by Age Group
The Perceived Mental Health by Age Group dataset spans from 2015 to 2022 and utilizes a consistent methodological framework to assess self-reported mental health. The data collection process involves standardized questionnaires distributed to a representative sample of Canadian adults across various age groups. This approach ensures that the data reflect a broad spectrum of the population, excluding those in full-time residence, on Indigenous reserves, or non-permanent residents, thus focusing on a generalizable sample of the Canadian populace.

### Perceived Mental Health by Gender and Sociodemographic Characteristics
The data for 2021 to 2023 from this dataset is sourced from the Canadian Community Health Survey (CCHS), which meticulously gathers data on mental health perceptions influenced by gender and other sociodemographic factors such as income, education, and employment. This survey selects participants through a stratified sampling design aimed at capturing diverse sociodemographic landscapes across Canada, enhancing the dataset's relevance for analyzing intersectional impacts on mental health.

## Detailed Measurement Discussion

Statistics Canada employs rigorous methodologies to ensure data accuracy and reliability. For example, the survey instruments are tested extensively before deployment to ensure they accurately capture the intended data. Additionally, data collectors are trained to adhere to strict guidelines to maintain consistency and reduce potential biases during data collection.

The sample for each dataset is designed to be representative of the Canadian population, stratified by key demographic variables to ensure comprehensive coverage across different regions and groups. This stratification helps in understanding specific sub-population dynamics within the broader national context. Data is then weighted to adjust for potential non-response biases and to represent the population accurately.

Despite these rigorous methods, the reliability of self-reported data can sometimes be challenging to assess due to the subjective nature of mental health perceptions. Statistics Canada addresses this by implementing multiple stages of data verification and validation, including automated data checks and manual reviews by experts. Furthermore, unusual or outlier responses undergo additional scrutiny to confirm their validity.



# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

```{=tex}
\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}
```
We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {.unnumbered}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References
